Combinatorial interpretations of holonomic recurrences

### 2021-07-25 article/main.tex
Describe lambda trees

\newcommand{\gf}[1]{\ensuremath{\mathcal{#1}}}
\newcommand{\enc}[1]{\ensuremath{\ols{#1}}}
\newcommand{\pointed}[1]{\ensuremath{{#1}^*}}
\newcommand{\weighted}[1]{\ensuremath{_{(#1)}}}
\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}

@@

\section*{Abstract}

@@

\section{Introduction}

\subsection{Notation}
\label{section-notation}

@@

\subsection{Structurally recursive objects}

@@

\subsection{Generating functions}

@@

\begin{example}
    \label{ex-bin-gf}
    Having our recursive definition of the object, we can easily construct a generating function for it. Let's create generating function that will count number of full binary trees consisting of exactly \(k\) internal nodes.

@@

\subsection{Holonomic recurrences}

@@

\begin{definition}
    \label{def-holo-1}
    \cite{holotoolkit}
    An infinite sequence \(A = (a_i)_{i=0}^{\infty}\) is holonomic (aka. P-finite, P-recursive or D-recursive) \(\textiff\) exists holonomic recurrence for it, ie.:

@@

\begin{definition}
    \label{def-holo-2}
    \cite{complexity}
    Equivalently to definition\ \ref{def-holo-1}, \(A = (a_i)_{i=0}^{\infty}\) is holonomic \(\textiff\) exists generating function in the following form:
    \[\sum_{i=0}^{r} p_i(z) \gf{A}^{(i)}(z) = b(z)\]
    for some fixed \(r \in \N\); \(b, p_0, \ldots, p_r \in \poly\) not all identically zero.
\end{definition}

Equivalence of these two definitions assures us that if we find generating function describing the desired objects satisfying definition\ \ref{def-holo-2}, we will be able to find holonomic recurrence mentioned in definition\ \ref{def-holo-1}.\cite{complexity}

\begin{example}
    \label{ex-gfun-rec}
    For generating function full binary trees defined in example\ \ref{ex-bin-gf}, ie.:
    \[\gf{B}(z) = 1 + z\gf{B}^2(z)\]
    we can use GFUN\cite{gfun} to immediately find holonomic recurrence of our interest:

@@

\subsection{Combinatorial interpretations}

@@

\subsection{Pre-order notation}

@@

\section{Known results}

\subsection{Binary trees}

@@

\subsection{Unary-binary trees}
\label{section-unary-binary}

@@

\begin{definition}
    \label{def-unary-binary-2}
    Left-leaning binary tree\todo{find official name; skewed binary tree is something different} is a binary tree with following constraint added: if a vertex has right child, it also have a left child.
\end{definition}

@@

To present the interpretation in bijective language, we will have to introduce several types of the the virtual leaves (you may want to revisit section \ref{section-notation} Notation):

\begin{enumerate}
    \item Vro - virtual leaf being right virtual-oneling;
    \item Vrt - virtual leaf being right virtual-twinling. We will divide this case into:
        \begin{enumerate}
            \item VrtUrv - Vri which uncle is a right virtual leaf;
            \item VrtUra - Vri which uncle is a right actual node;
            \item VrtUla - Vri which uncle is a left actual node.
        \end{enumerate}
    \item Vlt - virtual leaf left virtual-twinling. Note that from definition \ref{def-unary-binary-2}, left virtual leave can not have sibling being a actual node. We can divide also this case into:
        \begin{enumerate}
            \item VltUrv - Vlt which uncle is a right virtual leaf;
            \item VltUra - Vlt which uncle is a right actual node;
            \item VltUla - Vlt which uncle is a left actual node.
        \end{enumerate}
\end{enumerate}

@@

\subsection{Schröder trees}

@@

\section{Approaches}

\subsection{Introduction}

@@

\subsection{Defining the size carefully}

@@

\subsection{Tree binarization}

@@

As the most common trees are binary ones, it may be worth an effort to find the some structure based on binary trees that can represent original structure (see section~\ref{section-unary-binary}).

\subsection{Weights variablization}

@@

\subsection{Inductive translation}

@@

\subsection{Brutal iterative approach}

@@

As the main field of our interest was analyzis of terms of lambda trees, we will use it as an example to describe the method.\todo{Add definition}

At first let's create generating functions that will keep track of more than just number of vertices, but also counts of each kind of vertices in the tree. The generating function will be parameterized then using following arguments:
\begin{itemize}
    \item \(z\) counting total number of virtual leaves
    \item \(a\) counting number of application vertices
    \item \(l\) counting number of lambda abstraction vertices
    \item \(o\) counting number of bound variable vertices
    \item \(s\) counting number of successors of bound variable vertices
\end{itemize}

In order to make the image cleaner, we will drop the arguments of generating functions, ie. \(\gf{L} := \gf{L}(z, a, l, o, s)\)
\[\gf{L} = z a \gf{L}^2 + z l \gf{L} + \frac{o z}{1 - z s}\]
\todo{add some image}
\todo{describe how the frac is generated}

@@

\begin{lstlisting}
l^2*s^2*n*B(n) + ((4*a*o*s*u - 2*l^2*s - 2*l*s^2)*n + 6*a*o*s*u - 2*l^2*s - 3*l*s^2)*B(1 + n) + ((-4*a*o*u + l^2 + 4*l*s + s^2)*n - 8*a*o*u + 2*l^2 + 10*l*s + 3*s^2)*B(n + 2) + ((-2*l - 2*s)*n - 7*l - 8*s)*B(n + 3) + (5 + n)*B(n + 4)
\end{lstlisting}
\todo{rewrite to LaTeX}

You may notice that some potentially trivial terms are more complex that it seems.\todo{example}

@@

\section{Further work}

>>>>

\newcommand{\gf}[1]{\ensuremath{\mathcal{#1}}}
\newcommand{\enc}[1]{\ensuremath{\ols{#1}}}
\newcommand{\pointed}[1]{\ensuremath{{#1}^*}}
\newcommand{\weighted}[1]{\ensuremath{_{(#1)}}}
\newcommand{\size}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\LambdadB}{\ensuremath{\Lambda_{\text{dB}}}}

@@

\section*{Abstract}%
\label{sec:abstract}

@@

\section{Introduction}%
\label{sec:introduction}

\subsection{Notation}%
\label{sub:notation}

@@

\subsection{Structurally recursive objects}%
\label{sub:structurally_recursive_objects}

@@

\subsection{Generating functions}%
\label{sub:generating_functions}

@@

\begin{example}
    \label{ex:bin_gf}
    Having our recursive definition of the object, we can easily construct a generating function for it. Let's create generating function that will count number of full binary trees consisting of exactly \(k\) internal nodes.

@@

\subsection{Holonomic recurrences}%
\label{sub:holonomic_recurrences}

@@

\begin{definition}
    \label{def:holo_1}
    \cite{holotoolkit}
    An infinite sequence \(A = (a_i)_{i=0}^{\infty}\) is holonomic (aka. P-finite, P-recursive or D-recursive) \(\textiff\) exists holonomic recurrence for it, ie.:

@@

\begin{definition}
    \label{def:holo_2}
    \cite{complexity}
    Equivalently to definition\ \ref{def:holo_1}, \(A = (a_i)_{i=0}^{\infty}\) is holonomic \(\textiff\) exists generating function in the following form:
    \[\sum_{i=0}^{r} p_i(z) \gf{A}^{(i)}(z) = b(z)\]
    for some fixed \(r \in \N\); \(b, p_0, \ldots, p_r \in \poly\) not all identically zero.
\end{definition}

Equivalence of these two definitions assures us that if we find generating function describing the desired objects satisfying definition\ \ref{def:holo_2}, we will be able to find holonomic recurrence mentioned in definition\ \ref{def:holo_1}.\cite{complexity}

\begin{example}
    \label{ex:gfun-rec}
    For generating function full binary trees defined in example\ \ref{ex:bin_gf}, ie.:
    \[\gf{B}(z) = 1 + z\gf{B}^2(z)\]
    we can use GFUN\cite{gfun} to immediately find holonomic recurrence of our interest:

@@

\subsection{Combinatorial interpretations}%
\label{sub:combinatorial_interpretations}

@@

\subsection{Pre-order notation}%
\label{sub:pre_order_notation}

@@

\section{Known results}%
\label{sec:known_results}

\subsection{Binary trees}%
\label{sub:binary_trees}

@@

\subsection{Unary-binary trees}%
\label{sub:unary_binary_trees}

@@

\begin{definition}
    \label{def:unary_binary_2}
    Left-leaning binary tree\todo{find official name; skewed binary tree is something different} is a binary tree with following constraint added: if a vertex has right child, it also have a left child.
\end{definition}

@@

To present the interpretation in bijective language, we will have to introduce several types of the the virtual leaves (you may want to revisit section \ref{sub:notation} Notation):

\begin{enumerate}
    \item Vro - virtual leaf being right virtual-oneling;
    \item Vrt - virtual leaf being right virtual-twinling. We will divide this case into:
        \begin{enumerate}
            \item VrtUrv - Vri which uncle is a right virtual leaf;
            \item VrtUra - Vri which uncle is a right actual node;
            \item VrtUla - Vri which uncle is a left actual node.
        \end{enumerate}
    \item Vlt - virtual leaf left virtual-twinling. Note that from definition \ref{def:unary_binary_2}, left virtual leave can not have sibling being a actual node. We can divide also this case into:
        \begin{enumerate}
            \item VltUrv - Vlt which uncle is a right virtual leaf;
            \item VltUra - Vlt which uncle is a right actual node;
            \item VltUla - Vlt which uncle is a left actual node.
        \end{enumerate}
\end{enumerate}

@@

\subsection{Schröder trees}%
\label{sub:schröder_trees}

@@

\section{Approaches}%
\label{sec:approaches}

\subsection{Introduction}%
\label{sub:introduction}

@@

\subsection{Defining the size carefully}%
\label{sub:defining_the_size_carefully}

@@

\subsection{Tree binarization}%
\label{sub:tree_binarization}

@@

As the most common trees are binary ones, it may be worth an effort to find the some structure based on binary trees that can represent original structure (see section~\ref{sub:unary_binary_trees}).

\subsection{Weights variablization}%
\label{sub:weights_variablization}

@@

\subsection{Inductive translation}%
\label{sub:inductive_translation}

@@

\section{Brutal iterative approach}%
\label{sec:brutal_iterative_approach}

@@

As the main field of our interest was analyzis of terms of lambda trees in unary de Bruijn notation, we will use it as an example to describe the method.\todo{Add definition}

\subsection{Lambda terms}%
\label{sub:lambda_terms}

\begin{definition}
    Set of lambda terms \(\Lambda\) is defined recursively:
    \begin{itemize}
        \item if \(x\) is a variable, \(x \in \Lambda\)
        \item if \(M, N \in \Lambda\), then \((M N) \in \Lambda\) (application of function \(M\) to an argument \(N\))
        \item if \(x\) is a variable and \(M \in \Lambda\), then \((\lambda x . M) \in \Lambda\) (abstraction, ie.\ creating a function with an argument \(x\) and binding each occurence of \(x\) in \(M\) to this argument)
    \end{itemize}
\end{definition}

The definition above is quite popular, but introduces several problems. One of them is problem of potential collision of lambda-abstracted variables and difficulty of finding to which lambda abstraction is the variable is bound (or is it even free variable, ie.\ not bound). To address that, we can use de Bruijn indices.

\begin{definition}
    Set of lambda terms \(\LambdadB\) is defined recursively:
    \begin{itemize}
        \item if \(n \in \N\), \(n \in \LambdadB\) (\(n\) is a variable bound to lambda abstraction \(n\) levels above)
        \item if \(M, N \in \LambdadB\), then \((M N) \in \LambdadB\) (application)
        \item if \(M \in \LambdadB\), then \((\lambda M) \in \LambdadB\) (abstraction)
    \end{itemize}
\end{definition}

As you can see, we managed to eliminate the arbitrary variables. We can also easily check if given variable is bound by checking its depth.

\begin{example}
    \(\lambda x . \lambda y . x\) is equivalent to \(\lambda \lambda 1\)
\end{example}

\begin{example}
    \(\lambda x . (\lambda y . x) (\lambda y . y)\) is equivalent to \(\lambda (\lambda 1) (\lambda 0)\)
\end{example}

It also allows us to define tree representation of the terms.

\begin{definition}
    Injection from lambda terms in de Bruijn notation to binary trees \(\Phi\) is defined as follows:
    \begin{itemize}
        \item \(\Phi(0) = \n\; \no\; \no\)
        \item \(\Phi(n) = \n\; \no\; \Phi(n - 1)\)
        \item \(\Phi(\lambda M) = \n\; \Phi(M)\; \no\)
        \item \(\Phi(M N) = \n\; \Phi(M)\; \Phi(N)\)
    \end{itemize}
\end{definition}

To simplify the interpretation, we can introduce several kinds of the nodes: \(o\), \(s\), \(l\) and \(a\). They do not change the structure, but make the reading a bit easier:

\begin{definition}
    \(o\) is a leaf.
\end{definition}

\begin{definition}
    \(s\) is a node with just right child.
\end{definition}

\begin{definition}
    \(l\) is a node with just left child.
\end{definition}

\begin{definition}
    \(a\) is a node with both children.
\end{definition}

Which leads to following interpreation of \(\Phi\):

\begin{itemize}
    \item \(\Phi(0) = o\; \no\; \no\)
    \item \(\Phi(n) = s\; \no\; \Phi(n - 1)\)
    \item \(\Phi(\lambda M) = l\; \Phi(M)\; \no\)
    \item \(\Phi(M N) = a\; \Phi(M)\; \Phi(N)\)
\end{itemize}

You can note that not every binary tree represents valid term.

\begin{example}
    \(\n\; \no\; \n\; \n\; \no\; \no\; \no\) does not represent valid lambda term.
\end{example}

To see that, we can take a look at this tree with our node kinds used:
    \[s\; \no\; l\; o\; \no\; \no\; \no\]

\(s\) represents successor which can be applied only to \(n \in \N\), but is applied to lambda-abstracted term.

\subsection{Enhanced holonomic reccurence}%
\label{sub:enhanced_holonomic_reccurence}

As we have our structure defined, let's create generating functions that will keep track of more than just number of vertices, but also counts of each kind of vertices in the tree. The generating function will be parameterized then using following arguments:
\begin{itemize}
    \item \(z\) counting total number of virtual leaves
    \item \(o\) counting number of bound variable vertices
    \item \(l\) counting number of lambda abstraction vertices
    \item \(a\) counting number of application vertices
    \item \(s\) counting number of successors of bound variable vertices
\end{itemize}

In order to make the image cleaner, we will drop the arguments of generating functions, ie. \(\gf{L} := \gf{L}(z, o, l, a, s)\)
\[\gf{L} = z a \gf{L}^2 + z l \gf{L} + \frac{o z}{1 - z s}\]
\todo{add some image}
\todo{describe how the frac is generated}

@@

\[\begin{array}{rl}
        & (n - 4) l^2 s^2 L(n - 4)\\
        +& ((4 n - 10) a o s - (2 n - 6) l^2 s - (2 n - 5) l s^2) L(n - 3)\\
        +& ((-4 n + 8) a o + (n - 2) l^2 + (4 n - 6) l s + (n - 1) s^2) L(n - 2)\\
        +& ((-2 n + 1) l - 2 n s) L(n - 1)\\
        +& (n + 1) L(n)
\end{array}\]

You may notice that some potentially trivial terms are more complex that it seems. If we didn't count different types of vertices, we would achieve following holonomic recurrence:

\[\begin{array}{rl}
        & (n - 4) L(n - 4)\\
        +& L(n - 3)\\
        +& (2 n - 1) L(n - 2)\\
        +& (-4 n + 1) L(n - 1)\\
        +& (1 + n) L(n)
\end{array}\]

We can see that for example simple \(L(n-3)\) is expanded into \(((4 n - 10) a o s - (2 n - 6) l^2 s - (2 n - 5) l s^2) L(n - 3)\).

\subsection{The framework}%
\label{sub:the_framework}

@@

\section{Further work}%
\label{sec:further_work}
